---
title: "WNN vs Classical Algorithms Comparison"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{WNN vs Classical Algorithms Comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
Sys.setenv(LANG = "en")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(ReyCoquaisConstantin)
library(readxl)
library(dplyr)
library(ggplot2)
```

## 1. Data loading and preprocessing

```{r data_loading}
file_path <- system.file("extdata", "Elec-train.xlsx", package = "ReyCoquaisConstantin")

raw_data <- read_excel(file_path)

# 15 minutes timestep with daily periodicity --> 4*24 = 96
FREQ_DAILY <- 96
# our data starts at 01:15 am which is the 5th point of the day
elec_ts <- ts(raw_data$`Power (kW)`, frequency = FREQ_DAILY, start = c(1, 5))

# Calling an auxiliary logic to clean some outliers on the day before the day to be predicted
elec_cleaned_vec <- clean_elec_data(elec_ts)
elec_cleaned_ts  <- ts(elec_cleaned_vec, frequency = FREQ_DAILY, start = c(1, 5))

# removing the NA values in the end (the data to be predicted in fact)
idx_valid <- which(!is.na(elec_cleaned_ts))
last_idx <- max(idx_valid)

# Final series to train on
train_ts <- window(elec_cleaned_ts, end = time(elec_cleaned_ts)[last_idx])

plot(train_ts, main = "Cleaned Power Consumption", ylab = "Power (kW)",xlab="Days")
```

## 2. Fine-tuning K (Grid Search)

The paper presents the *Weighted Nearest Neighbors* implementation for time series forecasting, relying on three key parameters:

* **$w$ (Window Size):** The size of the sequence used for comparison.
* **$k$ (Number of Neighbors):** The number of most similar points to take into account when forecasting.
* **$h$ (Horizon):** The number of future values to predict.

Our series has a daily periodicity and we want to forecast one day of data. Therefore $w$ = $h$ = 96, 96 being the number of points in one day. We will focus on fine-tuning $k$ and we'll do this with a **Grid Search**. We are going to implement a **Rolling Origin** on the 7 last days of the series to then compute a mean of the MAPE and RMSE values over the 7 days for each $k$ of the Grid

```{r grid_search}
results <- data.frame(k = integer(), MAPE = numeric(), RMSE = numeric())
h_horizon <- 96 

# Testing k from 1 to 30 because WNN is really efficient
for (k in 1:30) {
  
  # Wrapper to call our wnn class
  wrapper_wnn <- function(train_set, h) {
    wnn_forecast(as.numeric(train_set), w = 96, h = h, k = k)
  }
  
  # Auxiliary cross validation
  scores <- cross_validate(train_ts, wrapper_wnn, h = h_horizon, n_folds = 7)
  
  results <- rbind(results, data.frame(k = k, MAPE = scores["MAPE"], RMSE = scores["RMSE"]))
}
```
Let's plot the RMSE evolution to choose the best $k$
```{r}
ggplot(results, aes(x = k, y = RMSE)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "darkblue", size = 2) +
  geom_point(data = results[which.min(results$RMSE), ], 
             color = "red", size = 4, shape = 1) +
  scale_x_continuous(breaks = 1:30) +
  labs(
    title = "Optimization of Hyperparameter k (Weighted Nearest Neighbors)",
    subtitle = paste0("Optimal k identified: ", results$k[which.min(results$RMSE)]),
    x = "Number of neighbors (k)",
    y = "Root Mean Squared Error (RMSE)"
  ) +
  theme_minimal()
```


```{r select_best}
# Selecting the best k
best_k <- results$k[which.min(results$RMSE)]
best_mape <- min(results$MAPE)
best_rmse <- min(results$RMSE)

cat(paste0("Best WNN model obtained with k = ", best_k, 
           "\n average MAPE  on validation window : ", round(best_mape, 3), "%",
            "\n average RMSE on validation window : ", round(best_rmse, 3), "%"))
```

## 3. Final forecast on 2/17/2010

We now implement our best model ($k=`r best_k`$) trained on the **whole** time series to predict the 2/17/2010.

```{r final_forecast}
final_forecast_values <- wnn_forecast(as.numeric(train_ts), w = 96, h = 96, k = best_k)

# Plotting
final_forecast_ts <- ts(final_forecast_values, frequency = 96, start = c(50, 1))

plot(train_ts, type='l', xlim=c(48, 51), main=paste("WNN forecast for Feb, 17th (k=", best_k,")"), ylab="kW")
lines(final_forecast_ts, col="red", lwd=2)
legend("topleft", legend=c("Historical", "Forecast WNN"), col=c("black", "red"), lty=1, lwd=1)
grid()
```

## 4. Comparison with the other methods implemented in inst/main.Rmd

In our previous study, we have come up with a *SARIMA(4,0,0)(0,1,1)[96]* being our best performer on the same sliding cross validation with a $MAPE$ = 4.07 and $RMSE$ = 14.8 kW

```{r comparison}
sarima_mape <- 4.07
sarima_rmse <- 14.82

comparison_df <- data.frame(
  Modèle = c("SARIMA (Part 1)", "WNN (R Package)"),
  `k optimal` = c("N/A", best_k),
  MAPE = c(sarima_mape, round(best_mape, 3)),
  RMSE = c(sarima_rmse, round(best_rmse, 3))
)

knitr::kable(comparison_df, caption = "Comaprative table of the models")
```

### Conclusion

The WNN model achieves a MAPE of **`r round(best_mape, 2)`%**, yielding performance metrics that are very close to the SARIMA model (**`r sarima_mape`%**). While slightly less precise, the WNN model stands out due to its radically superior computational efficiency.

In the context of high-frequency data ($f=96$), hyperparameter optimization for SARIMA models (and a fortiori the auto.arima procedure) becomes extremely computationally expensive. Conversely, the non-parametric WNN approach bypasses this complex estimation phase, offering quasi-instantaneous execution. Consequently, this model constitutes a particularly relevant alternative when deployment speed takes precedence over marginal gains in precision.
